Models of VOR learning in MHC knockout mice

### 2013-01-19 Notes/VORmodel.tex
comparison ov VOR training schemes
VOR: analytic discussion of multistate and two-state models

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{mydefs.tex} with its contents
%       or include mydefs.tex in zip/tar file
%    Replace \input{newsymb.tex} with its contents
%       or include newsymb.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figures/"}}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%equilibrium distribution
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\frg}{\W^\mathrm{F}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\F}{\boldsymbol{\Phi}}
\DeclareMathOperator{\SNR}{SNR}
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
\newcommand{\norm}{_0}
\newcommand{\inc}{_{\text{inc}}}
\newcommand{\dec}{_{\text{dec}}}
\newcommand{\incdec}{_{\text{inc/dec}}}
\newcommand{\wt}{_{\text{WT}}}
\newcommand{\kn}{_{\text{D$^\mathrm{b}$K$^\mathrm{b}$}}}
\newcommand{\Kn}{D$^\mathrm{b}$K$^\mathrm{b}$-/-}
\newcommand{\tpre}{t_{\text{pre}}}
% ----------------------------------------------------------------
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{Models of VOR learning in MHC knockout mice}
%
% Author List:
%
\author{Subhaneil Lahiri
%
}

@@

The untrained case will be described by $f\pot=f\pot\norm$.
Gain increase training  will be described by $f\pot=f\pot\inc<f\pot\norm$, and
gain decrease training  will be described by $f\pot=f\pot\dec<f\pot\norm$. Note that the forgetting matrix \eqref{eq:evolve} and the equilibrium distribution \eqref{eq:eqprob} depends on $f\pot$, which we will indicate with subscripts.

Before training, the synaptic distribution will in equilibrium with $f\pot\norm$.
During gain increase training, it will evolve according to \eqref{eq:evolve} with $f\pot\inc$:
%
\begin{equation}\label{eq:nopre}
  \pr(t) = \eq\norm \exp\prn{rt\frg\inc}.
\end{equation}
%
On the other hand, if the gain increase training follows gain decrease pre-training for some time period, $\tpre$:
%
\begin{equation}\label{eq:withpre}
  \pr(t) = \eq\norm \exp\prn{r\tpre\frg\dec} \exp\prn{r(t-\tpre)\frg\inc}.
\end{equation}
%

@@

\begin{table}
 \begin{center}
  \begin{tabular}{|l|c|c|c|c|}
    \hline
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    Model & \# states & pot,WT dep & \Kn\ dep & $\Delta f$\\
    \hline
    Cascade & 10 & $x=0.25$ & $x=0.3$ & -0.1 \\
    Multistate & 10 & $q=0.6$ & $q=0.8$ & -0.1 \\
    Two-state & 2 & $q=0.6$ & $q=0.8$ & -0.1 \\
    \hline
  \end{tabular}
 \end{center}
  \caption{Parameters used in simulations.}\label{tab:params}
\end{table}

>>>>

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{mydefs.tex} with its contents
%       or include mydefs.tex in zip/tar file
%    Replace \input{newsymb.tex} with its contents
%       or include newsymb.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figures/"}}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%equilibrium distribution
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\frg}{\W^\mathrm{F}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\F}{\boldsymbol{\Phi}}
\DeclareMathOperator{\SNR}{SNR}
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
\newcommand{\norm}{_0}
\newcommand{\inc}{_{\text{inc}}}
\newcommand{\dec}{_{\text{dec}}}
\newcommand{\incdec}{_{\text{inc/dec}}}
\newcommand{\wt}{_{\text{WT}}}
\newcommand{\kn}{_{\text{D$^\mathrm{b}$K$^\mathrm{b}$-/-}}}
\newcommand{\Kn}{D$^\mathrm{b}$K$^\mathrm{b}$-/-}
\newcommand{\tpre}{t_{\text{pre}}}
% ----------------------------------------------------------------
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{Models of VOR learning in MHC knockout mice}
%
% Author List:
%
\author{Subhaneil Lahiri
%
}

@@

The untrained case will be described by $f\pot=f\pot\norm$.
Gain-increase training  will be described by $f\pot=f\pot\inc<f\pot\norm$, and
gain-decrease training  will be described by $f\pot=f\pot\dec<f\pot\norm$. Note that the forgetting matrix \eqref{eq:evolve} and the equilibrium distribution \eqref{eq:eqprob} depends on $f\pot$, which we will indicate with subscripts.

Before training, the synaptic distribution will in equilibrium with $f\pot\norm$.
During gain-increase training, it will evolve according to \eqref{eq:evolve} with $f\pot\inc$:
%
\begin{equation}\label{eq:nopre}
  \pr(t) = \eq\norm \exp\prn{rt\frg\inc}.
\end{equation}
%
On the other hand, if the gain-increase training follows gain-decrease pre-training for some time period, $\tpre$:
%
\begin{equation}\label{eq:withpre}
  \pr(t) = \eq\norm \exp\prn{r\tpre\frg\dec} \exp\prn{r(t-\tpre)\frg\inc}.
\end{equation}
%

@@

\begin{table}
 \begin{center}
  \begin{tabular}{|l|c|c|c|c|}
    \hline
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    Model & \# states & pot,WT dep & \Kn\ dep & $\Delta f$\\
    \hline
    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 \\
    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.1 \\
    Two-state  & 2  & $q=0.6$  & $q=0.8$  & -0.1 \\
    \hline
  \end{tabular}
 \end{center}
  \caption{Parameters used in simulations.}\label{tab:params}
\end{table}

\subsection{Results}\label{sec:results}

The features of the actual experiments that we'd like to capture are:
\begin{itemize}
  \item Without pre-training, gain-increase learning is significantly faster in the wild-type than in the mutant.
  \item Gain-decrease pre-training is slightly faster in the wild-type, but not significantly so.
  \item After pre-training, gain-increase learning is significantly faster in the mutant than in the wild-type.
  \item For the wild-type, gain-increase learning is significantly faster without pre-training than with it.
  \item For the mutant, gain-increase learning is significantly faster with pre-training than without it.
\end{itemize}

We will try to gain some analytic insight to some of these models by looking at the slope of the learning curve at the start of gain-increase training.
This is proportional to the net-flux between the $\w=+1$ states and the $\w=-1$ states, measuring using the transition probabilities for gain-increase but the equilibrium distribution for either untrained or gain-decrease, assuming that pre-training lasts long enough to reach the equilibrium distribution for gain-decrease.

\subsubsection{Cascade}\label{sec:cascade}

\subsubsection{Multistate}\label{sec:multistate}

Consider the general uniform multistate model. 
Then the equilibrium distribution is given by
%
\begin{equation}\label{eq:mutltieq}
  \eq_i = \frac{1-\alpha}{1-\alpha^M}\,\alpha^{i-1},
  \qquad \text{where} \quad
  \alpha=\frac{f\pot q\pot}{f\dep q\dep}.
\end{equation}
%
If we take the limit $\alpha\rightarrow1$, this becomes $\frac{1}{M}$.

The net-flux from the $\w=+1$ states to the $\w=-1$ states is:
%
\begin{equation}\label{eq:multiflux}
  \Phi = \eq_{M/2+1}f'{}\dep q\dep - \eq_{M/2}f'{}\pot q\pot = \frac{1-\alpha}{1-\alpha^M}\,\alpha^{M/2-1}\prn{\alpha-\alpha'}f'{}\dep q\dep,
\end{equation}
%
where primed values correspond to the new value of $f\pot$.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:multiWTnopre}
  \Phi = -\frac{2\Delta fq}{M},
\end{equation}
%
where it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:multiWTpre}
\begin{aligned}
  \Phi &= 16(\Delta f)^2q \, \frac{(1-2\Delta f)^{M/2-1} (1+2\Delta f)^{M/2-1}}
          {(1-2\Delta f)^M - (1+2\Delta f)^M} \\
       &= -\frac{4\Delta fq}{M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%
So, we see that pre-training will speed up learning when $\Delta f$ is small.
On the other hand, if $\Delta f$ is close to $-\half$, pre-training will initially slow down learning a lot.

Now, consider the mutant, for which $q\pot=\beta q\dep=q$, $\beta<1$.
Without pre-training:
%
\begin{equation}\label{eq:multiKNnopre}
  \Phi = -\frac{2(1-\beta)\beta^{M/2-1}\Delta fq}{1-\beta^M}.
\end{equation}
%
With pre-training:
%
\begin{equation}\label{eq:multiKnpre}
\begin{aligned}
  \Phi &= -4\Delta f q \, \frac{(1-2\Delta f) - \beta(1+2\Delta f)}
          {(1-2\Delta f)^M - \beta^M(1+2\Delta f)^M}   \,
          \beta^{M/2-1}(1-2\Delta f)^{M/2-1} (1+2\Delta f)^{M/2-1}.
\end{aligned}
\end{equation}
%

\subsubsection{Two-state}\label{sec:binary}

This model can be solved exactly:
%
\begin{multline}\label{eq:binarysol}
  \eq = \frac{(f\dep q\dep, f\pot q\pot)}{\lambda},
  \qquad
  \pr(t) = \eq + (\pr(t)-\eq)\e^{-\lambda rt},\\
  \qquad \text{where} \quad
  \lambda = f\pot q\pot + f\dep q\dep.
\end{multline}
%
But it is easier to just substitute $M=2$ into the formulae in \sref{sec:multistate}.
In this case, the initial rate of change encapsulates the whole solution, as there is only a single exponential decay.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:binWTnopre}
  \Phi = -\Delta f q,
\end{equation}
%
where it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:binWTpre}
\begin{aligned}
  \Phi &= -2\Delta f q.
\end{aligned}
\end{equation}
%
So, we see that pre-training will always speed up learning.

Now, consider the mutant, for which $q\pot=\beta q\dep=q$, $\beta<1$.
Without pre-training:
%
\begin{equation}\label{eq:binKNnopre}
  \Phi = -\frac{2\Delta f q}{1+\beta},
\end{equation}
%
which is always larger that \eqref{eq:binWTnopre}.
With pre-training:
%
\begin{equation}\label{eq:binKnpre}
\begin{aligned}
  \Phi &= -4\Delta f q \, \frac{(1-2\Delta f) - \beta(1+2\Delta f)}
          {(1-2\Delta f)^2 - \beta^2(1+2\Delta f)^2}.
\end{aligned}
\end{equation}
%
